---
title: "Environmental data sort"
output: html_notebook
---


Sorting out the envirnonmental layers to get a raster with 100m resolution grid cells right up to the coastline. Going to use both mode and target landcover classes. 


```{r, packages}

library(raster)

```


## land cover map

```{r lcm}

# load in 100m perc landcover
lcm <- raster::stack('C:/Users/thoval/OneDrive - UKCEH/Documents/DECIDE/DECIDE_WP1/data/raw_data/environmental/CEH_land_cov/lcm_2015/25m_raster/GB/data/lcm2015gb100perc.tif')

names(lcm) <-  c("sea",
                 "broad_wood",
                 "conif_wood",
                 "arable",
                 "impr_grass",
                 "neutr_grass",
                 "calc_grass",
                 "acid_grass",
                 "fen_marsh_swamp",
                 "heather",
                 "heath_grass",
                 "bog",
                 "inland_rock",
                 "saltwater",
                 "freshwater",
                 "sup_lit_rock",
                 "sup_lit_sed",
                 "lit_rock",
                 "lit_sed",
                 "saltmarsh",
                 "urban",
                 "suburban")

# remove sea
lcm <- dropLayer(lcm, 1)


# plot(lcm)


```

## climate data load and create bioclim variables

2010-2019 inclusive.

```{r climate_bioclim, warning = F, message=F}


# load in climate
path <- "C:/Users/thoval/OneDrive - UKCEH/Documents/DECIDE/DECIDE_WP1/data/raw_data/environmental/HadUK_dat/"

vars <- c("rainfall", "tasmax", "tasmin")

out_var <- list()

for(i in vars) {
  
  # list all the files with a given structure
  files <- list.files(pattern = i, path = path, full.names = T)
  
  # initialise list
  st_l <- list()
  
  for(j in 1:length(files)){
    
    # get the files for each year
    # need to use stack so that each month is a different layer
    env_d <- raster::stack(files[j])
    
    # rename layers 1:12 - months of the year
    names(env_d) <- paste("M", seq(1:12), sep = "_")  
    
    # store as an object to use as an index to average across later
    ind <- names(env_d)
    
    # store the object in the list to use later
    st_l[[j]] <- env_d
    
    
  }
  
  print(i)
  
  # now want to average the same layer across multiple raster stacks
  # basically because I want to get the average min, max temp and rainfall 
  # for each month over 10 years, 2010 to 2019 inclusive
  st_o <- stackApply(stack(st_l), # creates a stack of all objects in the list
                     fun = mean, # takes the mean
                     indices = ind, # provide an index to get the names
                     na.rm = T)
  out_var[[i]] <- st_o
  
}

# convert to the 19 bioclimatic variables
had_bv <- dismo::biovars(prec = out_var$rainfall, tmin = out_var$tasmin, tmax = out_var$tasmax)

# name real names
names(had_bv) <- c("AnnualTemp",
                   "MeanDiRange",
                   "Isotherm",
                   "TempSeasonality",
                   "MaxTempWarmestMonth",
                   "MinTempColdestMonth",
                   "TempAnnualRange",
                   "MeanTempWetQuarter",
                   "MeanTempDriestQuarter",
                   "MeanTempWarmQuarter",
                   "MeanTempColdQuarter",
                   "AnnualPrecip",
                   "PrecipWetMonth",
                   "PrecipDriestMonth",
                   "PrecipSeasonality",
                   "PrecipWettestQuarter",
                   "PrecipDriestQuarter",
                   "PrecipWarmQuarter",
                   "PrecipColdQuarter")

# plot(had_bv)

```

## Remove collinear weather variables

Probably better to do PCA reduction 

```{r weath_colliner}

v <- usdm::vifcor(had_bv)
had_bv_nocorrs <- raster::dropLayer(had_bv, match(names(had_bv), v@excluded))

names(had_bv_nocorrs)

```

## elevation, slope and aspect

Elevation was created and reprojected on Datalabs as it was too large to do on own PC. The script is in DWPTWO project in notebook "processing_elev_raster.R". So, for now, just load in the data that i transferred back from Datalabs - should be on the same scale.

A copy of the code is in the chunk below; I wasn't careful about absolute file paths etc.

```{r load_elev}

elev_100 <- raster('../../Data/environmental_data/elevation_map_coasts_fixed_GB_100m.tif')


```


```{r, elevation}



# # unzip(zipfile = '/data/notebooks/rstudio-vizz/data/elevation/Copernicus_Elevation.zip', 
# #       exdir = '/data/notebooks/rstudio-vizz/data/elevation/')
# 
# library(raster)
# 
# #####    Elevation data
# 
# # just a simple walkthrough of the elevation data that's available
# # the files are pretty big though so they will need to be cropped before using.
# 
# # the elevation data are found here:
# list.files('/data/notebooks/rstudio-vizz/data/elevation/Copernicus_Elevation/')
# 
# # this contains a 100m raster for the UK combined
# uk_100m <- raster::stack('/data/notebooks/rstudio-vizz/data/elevation/Copernicus_Elevation/elevation_UK.tif')
# uk_100m
# 
# 
# # there are also rasters available at 25m
# el1 <- raster("/data/notebooks/rstudio-vizz/data/elevation/Copernicus_Elevation/eu_dem_v11_E30N30/eu_dem_v11_E30N30.TIF")
# el2 <- raster("/data/notebooks/rstudio-vizz/data/elevation/Copernicus_Elevation/eu_dem_v11_E30N40/eu_dem_v11_E30N40.TIF")
# 
# # merge the rasters together
# elev_100 <- merge(el1, el2)
# 
# # remove iceland
# ext_proj <- extent(matrix(c(3000000,3000000, 3900000, 4500000), ncol = 2))
# elev_100_crop <- crop(elev_100, ext_proj) 
# plot(elev_100_crop)
# 
# 
# # landcover layer for reprojecting to:
# lcm <- raster("/data/notebooks/rstudio-setupconsthomas/DECIDE_constraintlayers/Data/environmental_data/100mRastOneLayer.grd")
# 
# elev_proj <- raster::projectRaster(elev_100_crop, lcm)
# elev_proj
# plot(elev_proj)
# 
# ## crop to GB
# # download map GB
# uk_map <- st_as_sf(getData("GADM", country = "GBR", level = 1, path='Data/environmental_data'))
# uk_map <- st_transform(uk_map, 27700)
# 
# # remove nrothern ireland
# gb_map <- uk_map[uk_map$NAME_1 != 'Northern Ireland',]
# 
# # check
# plot(st_geometry(gb_map))
# 
# # convert to spatial for use in raster::mask()
# gb_mask <- as_Spatial(gb_map)
# gb_mask
# 
# # mask elevation
# m_gb <- raster::mask(elev_proj, gb_mask[1])
# 
# plot(m_gb, xlim = c(330000, 360000), ylim = c(440000, 470000))
# 
# writeRaster(m_gb, filename = 'Data/environmental_data/elevation_map_coasts_fixed_GB_100m.tiff')



```



## get all variables on scale of land cover map

Using bilinear interpolation

```{r reproj}

rm(list=setdiff(ls(), c("had_bv_nocorrs", "lcm", "elev_100")))

clim_reproj <- projectRaster(had_bv_nocorrs, lcm[[1]], method = 'bilinear')
plot(clim_reproj[[1]]) ## need to remove Northern Ireland

```

## Get to climate to GB only


```{r}


## crop to GB
# download map GB
library(sf)
uk_map <- st_as_sf(getData("GADM", country = "GBR", level = 1, path='../../data/environmental_data'))
uk_map <- st_transform(uk_map, 27700)

# remove nrothern ireland
gb_map <- uk_map[uk_map$NAME_1 != 'Northern Ireland',]

# check
plot(st_geometry(gb_map))

# convert to spatial for use in raster::mask()
gb_mask <- as_Spatial(gb_map)
gb_mask

# mask elevation
clim_reproj_gb <- raster::mask(clim_reproj, gb_mask[1])
plot(clim_reproj_gb[[1]])


```


## stack all together

```{r stack}

env_dat <- raster::stack(lcm, clim_reproj_gb, elev_100)
env_dat
names(env_dat)

# check
plot(env_dat[[31]], xlim = c(340000, 350000), ylim = c(679000, 690000)) 
plot(env_dat[[23]], xlim = c(340000, 350000), ylim = c(679000, 690000)) 
plot(env_dat[[24]], xlim = c(340000, 350000), ylim = c(679000, 690000)) 
plot(env_dat[[25]], xlim = c(340000, 350000), ylim = c(679000, 690000)) 

```


## Slope and aspect

```{r slope_asp}

slope <- terrain(env_dat[[31]], opt = "slope", unit = "degrees")
aspect <- terrain(env_dat[[31]], opt = "aspect", unit = "degrees")

par(mfrow = c(1,2))
plot(slope)
plot(aspect)
par(mfrow = c(1,1))

env_dat <- stack(env_dat, slope, aspect)

```


## check that the rasters follow the coast in aberlady

```{r, check}

## check
library(ggmap)

register_google(key = "AIzaSyDzfFo6Jq1g65BauVhLL-SexOYS5OuH-kA")
location = c(-2.860564, 56.014902) # aberlady
gmap <- get_map(location = location, zoom = 12)
ablad <- crop(env_dat[[23]], extent(c(340000, 350000, 679000, 690000)))
ogd <- projectRaster(ablad, crs="+proj=longlat +datum=WGS84") %>% as.data.frame(xy=T)
colnames(ogd) <- c('x','y','z')

ggmap(gmap) +
  geom_tile(data = ogd, aes(x=x,y=y, fill = z), alpha = 0.7) +
  coord_quickmap() +
  scale_fill_gradient(high = 'yellow', low = 'darkblue', na.value = NA) +
  labs(x= '', y = '')

```


All good :)

## Write to file

```{r}

# # Write to file
# writeRaster(env_dat, filename = '../../data/environmental_data/envdata_fixedcoasts_nocorrs_100m_GB.grd',
#             format = 'raster')

```


## Check to make sure can read in

(Had problem with last one)

```{r correlated conds}

ed <- raster::stack('../../data/environmental_data/envdata_fixedcoasts_nocorrs_100m_GB.grd')
ed

plot(ed[[32]])
plot(ed[[33]])

```

All good!! 

