---
title: "Pseudoabsence exploration moths"
author: "Thomas MM"
date: "16/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)


```

## Exploring list numbers with moth data

In order to get the most accurate information about pseudoabsences for each species, we need to look at the list lengths. Lists will be defined as the sightings from the same date at the same location - a proxy for someone submitting their sightings (assumes no two different people are in the same 100m location on a given day). This is because the data contain sightings of moths found during day (e.g. from people out walking), night (moth traps) and some that are found in both (species that are found in both daytime walks and in moth traps, e.g mother shipton and burnet companion), hereafter 'day-and-night'.

The problem lies with this final group and creating pseudoabsences. Comparing sightings of a true day-flying moth to pseudoabsences generated using moths that were actually flying by night would mean that we'd get a pseudpabsence from a location and time that the moth of interest would never have been found at. 

# A naive set of questions and some attempted answers: 

Does this matter because given that a day-and-night flying moth is likely to be active in the same broad geographic region during the day and night? This means that we won't be biasing through differences in habitat. I guess it matters if the locations where day-and-night flyers come out are different during the day and night. I.e. if moth mother shipton comes out on grasslands during the day and only wood at night. I'm not too sure whether this ever happens.

If day-and-night species are more common in the datasets (which they appear to be from the results below) then this means there are likely to be relatively more pseudoabsences generated by them compared to true day-fliers only. I am not too sure how this could bias the distribution estimates. Also, we are already removing "repeats" by taking out sightings of the same species, on the same day from the same location so this would "thin out" the influence of loads of a day-and-night flying species being caught in the same moth trap (although it wouldn't if there were loads of the same day-and-night species caught in the same trap).

This could also matter because the catching mechanism for night moths lures individuals in, whereas not true in the day. If the "catching radius" of the moth trap is smaller than the resolution we're interested in looking at for the environmental variables (100m), then it shouldn't matter (I think). 

## what do we expect?

Day-flying moth lists are likely to be shorter than for night-flying moths, because in the latter you expect to get many species being lured to the light source, whereas in the day time it will be incidental sightings only. So, if we create histograms of the length of lists for each species we can compare daytime, nighttime and day-and-night species. We would expect most day-time moths to be found on short lists, most night-time moths to be found on long lists and day-and-night moths to (maybe) have a bimodal distribution.

## Why do this?

If there is a bimodal distribution for day-and-night moths then we could determine a threshold for length of list above which were remove all the sighitngs of day-and-night. This way we can make sure that we are analysing records taken from species sighted in comparable ways. Alternatively, we could remove all species that have bimodal distributions to make 100% sure that the species we are comparing to are day-flying.

# the data

These data have the duplicates removed. So there are no reports of the same species from the same location on the same day.

```{r moths, echo = F, message=FALSE, warning=FALSE}

# Load the data

spdf <-  read_csv("Data/moth_no_duplicates.csv")

```

# step 1

Get a list of all day-flying species from the excel sheet from Marc B.

```{r day_fliers}

## get day-flying moths
dfm <- c("Jordanita globulariae",
         "Adscita statices",
         "Adscita geryon",
         "Zygaena purpuralis",
         "Zygaena loti",
         "Zygaena exulans",
         "Zygaena viciae",
         "Zygaena filipendulae",
         "Zygaena lonicerae",
         "Zygaena lonicerae/trifolii",
         "Zygaena trifolii",
         "Lasiocampa quercus",
         "Hemaris tityus",
         "Hemaris fuciformis",
         "Idaea muricata",
         "Scotopteryx bipunctaria",
         "Epirrhoe tristata",
         "Minoa murinata",
         "Rheumaptera hastata",
         "Odezia atrata",
         "Perizoma albulata",
         "Eupithecia pygmaeta",
         "Archiearis parthenias",
         "Archiearis notha",
         "Macaria carbonaria",
         "Macaria brunneata",
         "Chiasmia clathrata",
         "Pseudopanthera macularia",
         "Lycia lapponaria",
         "Lycia zonaria",
         "Ematurga atomaria",
         "Glacies coracina",
         "Siona lineata",
         "Perconia strigillaria",
         "Orgyia antiqua",
         "Orgyia recens",
         "Parasemia plantaginis",
         "Tyria jacobaeae",
         "Phytometra viridaria",
         "Euclidia glyphica",
         "Euclidia mi",
         "Tyta luctuosa",
         "Panemeria tenebrata",
         "Heliothis viriplaca",
         "Heliothis maritima",
         "Eremobia ochroleuca",
         "Photedes captiuncula",
         "Cerapteryx graminis",
         "Anarta melanopa",
         "Anarta myrtilli",
         "Anarta cordigera")

dfm

```


# step 2

Create an id for visit (vis_id) which combines date, lat and lon. This way we can identify sightings from the same day and location, i.e. visit. 

Create a factor of whether a moth is a day or night flying species so that I can see which to look for when plotting later. 


```{r visits}

length(unique(spdf$sp_n))
length(spdf$sp_n %in% dfm)

spdf <- spdf %>% mutate(vis_id = paste(date, lat, lon, sep = "_"), 
                        flyer = ifelse(spdf$sp_n %in% dfm, "day", "night"))


# add a column to make sure species names are visible in the gridded plots below
var_width = 15
new_n <- data.frame(o_n = unique(spdf$sp_n), nn =str_wrap(unique(spdf$sp_n), width = var_width))
# head(new_n)


```


# step 3

Now need the length of the lists that each sighting is a part of and plot the number by species. This is the total number of individuals caught per visit (date and location only), NOT the number of different species.

I'm only plotting a subset of the data for night-flyers because there are so many. Important to note that the y-scale varies between each plot.


```{r list_length}


# for each visit ID, get the number of visits made 
spdf <- spdf %>% group_by(vis_id) %>% # group by the visit ID
  mutate(n_lists = length(vis_id), # find the number of lists within each of those visits
         n_spp = length(unique(sp_n))) %>% # find the unique number of species found within each list
  ungroup %>% 
  mutate(b_n = new_n$nn[match(sp_n, new_n$o_n)]) # match the nicer-looking names 

# hist(spdf$n_lists)

```


```{r plotting2, echo = F}

day_sub <- unique(spdf$sp_n[spdf$sp_n %in% dfm])

subset(spdf, sp_n %in% day_sub[1:20]) %>% 
  ggplot() +
  geom_histogram(aes(x = n_spp, fill = flyer), bins = 30) +
  facet_wrap(~b_n, scales = "free") +
  theme_bw()  +
  xlab("number of lists")

subset(spdf, sp_n %in% day_sub[21:40]) %>% 
  ggplot() +
  geom_histogram(aes(x = n_spp, fill = flyer), bins = 30) +
  facet_wrap(~b_n, scales = "free") +
  theme_bw() +
  xlab("number of lists")

subset(spdf, sp_n %in% day_sub[41:46]) %>% 
  ggplot() +
  geom_histogram(aes(x = n_spp, fill = flyer), bins = 30) +
  facet_wrap(~b_n, scales = "free") +
  theme_bw() +
  xlab("number of lists")


# The number of day fliers takes a lot longer because there are many more species
`%!in%` = Negate(`%in%`)

night_sub <- unique(spdf$sp_n[spdf$sp_n %!in% dfm])

subset(spdf, sp_n %in% night_sub[1:20]) %>% 
  ggplot() +
  geom_histogram(aes(x = n_spp), fill = 'darkblue', bins = 30)+
  facet_wrap(~b_n, scales = "free") +
  theme_bw() +
  xlab("number of lists")


```


There are far fewer sightings of day-flyers than night-flyers (I guess as expected). There are several species of day-fliers that have similar list distributions to those of night-fliers. These also appear to be the day-fliers with the largest number of sightings (although not always, e.g. Zygaena filipendulae has a lot of sightings but doesn't have a 'tail').

## step 4

Now we want to plot the unique number of species per visit, should be the same as the plot above because I removed duplicates from the original file.

```{r plotting, echo = F}

day_sub <- unique(spdf$sp_n[spdf$sp_n %in% dfm])

subset(spdf, sp_n %in% day_sub[1:20]) %>% 
  ggplot() +
  geom_histogram(aes(x = n_lists, fill = flyer), bins = 30) +
  facet_wrap(~b_n, scales = "free") +
  theme_bw()  +
  xlab("number of lists")

subset(spdf, sp_n %in% day_sub[21:40]) %>% 
  ggplot() +
  geom_histogram(aes(x = n_lists, fill = flyer), bins = 30) +
  facet_wrap(~b_n, scales = "free") +
  theme_bw() +
  xlab("number of lists")

subset(spdf, sp_n %in% day_sub[41:46]) %>% 
  ggplot() +
  geom_histogram(aes(x = n_lists, fill = flyer), bins = 30) +
  facet_wrap(~b_n, scales = "free") +
  theme_bw() +
  xlab("number of lists")


# The number of day fliers takes a lot longer because there are many more species
`%!in%` = Negate(`%in%`)

night_sub <- unique(spdf$sp_n[spdf$sp_n %!in% dfm])

subset(spdf, sp_n %in% night_sub[1:20]) %>% 
  ggplot() +
  geom_histogram(aes(x = n_lists), fill = 'darkblue', bins = 30)+
  facet_wrap(~b_n, scales = "free") +
  theme_bw() +
  xlab("number of lists")


```