---
title: "Sorting out species data"
output: html_notebook
---

The 'no duplicates' files contain no duplicates of the same species on the same day; but do contain duplicates on different days.

```{r}

knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(BRCmap)
library(lubridate)

# do we want to write the CSVs out?
Write = TRUE


```


This script is to create the datasets for use later on. Wanted to keep the scripts for all taxa in one place for ease of access. This script creates datasets with the raw data at 100m resolution and removes duplicates of same species, in the same place on the same date. In the final pseudoabsence generation we remove any duplicates in the same grid cell; this is done on Lotus during the pseudoabsence scripts.


## Moths


```{r load_files, message=F, warning=F}

# get the files from the NRMS and bind them together
f <- list.files("../../data/raw_data/moth/", full.names = T)
f

out_dat <- list()

for(i in grep(pattern = "NMRS", x = f, value = T)){
  print(i)
  
  out_dat[[i]] <- read_csv(file = i)
  
}

main_df <- do.call("rbind", out_dat)

head(main_df)


# get the species name
l_n <- read_csv(file = grep(pattern = "lepidoptera_names", x = f, value = T))


```


choose 100m resolution, match latin names.

```{r, data_faff}

# only get precision 100 and recorded on the same date
df <- subset(main_df, DT_ID == 1  & # start and end on same day
                 TO_PRECISION == 100) %>% 
  mutate(date = dmy(TO_STARTDATE),
         yr = year(date),
         jd = yday(date),
         sp_n = l_n$NAME[match(CONCEPT, l_n$CONCEPT)],
         com_n = l_n$NAME_ENGLISH[match(CONCEPT, l_n$CONCEPT)],
         ig = l_n$INFORMAL_GROUP[match(CONCEPT, l_n$CONCEPT)],
         ag = l_n$ADDITIONAL_GROUP[match(CONCEPT, l_n$CONCEPT)])


```



```{r, day_fliers}

# day fliers as based on the rmarkdown document and 
# the expert decision makers
dfm <- c("Jordanita globulariae",
         "Adscita statices",
         "Adscita geryon",
         "Zygaena purpuralis",
         "Zygaena loti",
         "Zygaena exulans",
         "Zygaena viciae",
         "Zygaena filipendulae",
         "Zygaena lonicerae",
         "Zygaena lonicerae/trifolii",
         "Zygaena trifolii",
         # "Lasiocampa quercus",
         "Hemaris tityus",
         "Hemaris fuciformis",
         "Idaea muricata",
         "Scotopteryx bipunctaria",
         "Epirrhoe tristata",
         "Minoa murinata",
         "Rheumaptera hastata",
         "Odezia atrata",
         "Perizoma albulata",
         "Eupithecia pygmaeta",
         "Archiearis parthenias",
         "Archiearis notha",
         "Macaria carbonaria",
         "Macaria brunneata",
         "Chiasmia clathrata",
         "Pseudopanthera macularia",
         "Lycia lapponaria",
         "Lycia zonaria",
         "Ematurga atomaria",
         "Glacies coracina",
         "Siona lineata",
         "Perconia strigillaria",
         "Orgyia antiqua",
         "Orgyia recens",
         "Parasemia plantaginis",
         "Tyria jacobaeae",
         "Phytometra viridaria",
         "Euclidia glyphica",
         "Euclidia mi",
         "Tyta luctuosa",
         "Panemeria tenebrata",
         "Heliothis viriplaca",
         "Heliothis maritima",
         # "Eremobia ochroleuca",
         "Photedes captiuncula",
         # "Cerapteryx graminis",
         "Anarta melanopa",
         "Anarta myrtilli",
         "Anarta cordigera")

dfm_df <- df[df$sp_n %in% dfm,]

```

Get eastings northings and bind with data frame (rename lon lat for simplicity)

```{r, east_norths}

# lat lon from grid refs
ll <- gr_let2num(dfm_df$TO_GRIDREF, centre = TRUE)
head(ll)

dfm_df <- dfm_df %>% 
  mutate(lon = ll$EASTING,
         lat = ll$NORTHING,
         id = paste(sp_n, date, lon, lat, sep = "_")) ## create an id column for identifying duplicates of the same species from the same location on the same date)

head(dfm_df)

if(Write){
  
  write.csv(dfm_df, file = '../../data/edited_insect_data/moth/DayFlyingMoths_EastNorths_with_duplicates.csv')
  
}



```


Remove duplicates of same species, on same day in same location


```{r, remove_dups}

dfm_df_no_dups <- dfm_df %>% 
  distinct(sp_n, date, lon, lat, .keep_all = T) # use these columns to determine whether or not distinct

if(Write){
  
  write.csv(dfm_df_no_dups, file = '../../data/edited_insect_data/moth/DayFlyingMoths_EastNorths_no_duplicates.csv')
  
}

```


## Butterflies

```{r, restart}

# remove everything except 'write'
rm(list=setdiff(ls(), "Write"))


```


```{r, butt_files, message=FALSE}


# get list of files
f <- list.files(path = '../../data/raw_data/butterfly/', full.names = TRUE)
f

# read subset of NBN data to test 2000 - 2010
b00 <- read_csv(f[1])
b10 <- read_csv(f[2])

b00 <- rbind(b00, b10)

# get species codes
l_n <- read_csv(f[3])


```


choose 100m resolution, match latin names, east norths

```{r, data_faffing}


b002 <- subset(b00, DT_ID == 1  & 
                 TO_PRECISION == 100) %>% 
  mutate(date = dmy(TO_STARTDATE),
         yr = year(date),
         jd = yday(date),
         sp_n = l_n$NAME[match(CONCEPT, l_n$CONCEPT)],
         com_n = l_n$NAME_ENGLISH[match(CONCEPT, l_n$CONCEPT)],
         ig = l_n$INFORMAL_GROUP[match(CONCEPT, l_n$CONCEPT)],
         ag = l_n$ADDITIONAL_GROUP[match(CONCEPT, l_n$CONCEPT)])


# east norths from grid refs
ll <- gr_let2num(b002$TO_GRIDREF, centre = TRUE)

# add eastings northings as lon lat
b_df <- b002 %>% 
  mutate(lon = ll$EASTING,
         lat = ll$NORTHING, 
         id = paste(sp_n, date, lon, lat, sep = "_")) ## create an id column for identifying duplicates of the same species from the same location on the same date

head(b_df)

if(Write){
  
  write.csv(b_df, file = '../../data/edited_insect_data/butterfly/butterfly_EastNorths_with_duplicates.csv')
  
}



```


```{r, butterfly_remove_duplicates}


b_df_no_dups <- b_df %>% 
  distinct(sp_n, date, lon, lat, .keep_all = T) # use these columns to determine whether or not distinct

if(Write){
  
  write.csv(b_df_no_dups, file = '../../data/edited_insect_data/butterfly/butterfly_EastNorths_no_duplicates.csv')
  
}


```





