---
title: "Making effort layers"
output: html_notebook
---


```{r}

library(raster)
library(tidyverse)

Write = TRUE

```


For the simulation work, we need to make an effort layer of recording activities for moths and butterflies. This is going to be the number of unique days with a sighting of any species per grid cell.

## moths

```{r, load_data}

# moth
mdf <- read_csv('../../data/edited_insect_data/moth/DayFlyingMoths_EastNorths_no_duplicates.csv')

# butterflies
bdf <- read_csv('../../data/edited_insect_data/butterfly/butterfly_EastNorths_no_duplicates.csv')

# 100m raster
env <- raster::stack('../../data/environmental_data/edat_nocorrs_nosea_cropped.grd')
env <- subset(env, "elevation_UK")

par(mfrow = c(1,2))
plot(env, main = 'moths')
points(x = mdf$lon, y = mdf$lat, pch = 20, cex = 0.1)


plot(env, main = 'butterflies')
points(x = bdf$lon, y = bdf$lat, pch = 20, cex = 0.1)

par(mfrow = c(1,1))


```

## aggregate raster to desired resolution

```{r, agg_resolution}

km1 <- raster::aggregate(env, 10)
km10 <- raster::aggregate(env, 100)

```
```{r, num_cells}


cbind(# 1km GB raster number of cells
  raster::as.data.frame(km1) %>% 
    na.omit() %>% 
    summarise(number_cells_1km_GB = length(layer)),
  
  # 10km GB raster number of cells
  raster::as.data.frame(km10) %>% 
    na.omit() %>% 
    summarise(number_cells_10km_GB = length(layer)))



```


## Changing grouping level

Get the number of visits to a grid cell, but only allow visits on different days. 

```{r}

mdf_no_spp <- mdf  %>% 
  distinct(lon, lat, date)


bdf_no_spp <- bdf  %>% 
  distinct(lon, lat, date)


```



## effort function

Creating an effort function that takes a coordinates matrix, a raster and an aggregation factor (and an optional naming of the output layer). It converts the raster to 0s, counts the number of points that fall in each cell and stores them in the raster. It then applies a mask to the raster to force NAs into the areas falling outside the mask. If the 'fact' argument = 1 then no change is made to the cell size of the raster itself. If it is > 1 then it aggregates by that factor. 

```{r, effort_function}

get_effort <- function(coords_df, # two column matrix of (lon, lat)
                       rast, # raster for initial filling
                       fact, # factor to increase spatial resolution of raster - relative to initial resolution
                       name = NULL) # name to give layer
{  
  
  
  if(fact < 1) stop('aggregation factor must be > 0')
  if(fact > 1) { message('###   aggregating   ###'); rast <- raster::aggregate(rast, fact = fact) }
  
  # convert raster to 0s
  r <- rast
  r[] <- 0
  
  # get counts per cell
  counts <- table(cellFromXY(r, coords_df))
  
  # fill matrix
  r[as.numeric(names(counts))] = counts
  
  message('###   masking   ###')
  agg_mask <- raster::mask(r, rast)
  names(agg_mask) <- name
  
  return(agg_mask)
}



```

## moths 1 + 10km

```{r, moths_1+10km}

mths_1km <- get_effort(coords_df = cbind(mdf_no_spp$lon, mdf_no_spp$lat),
                       rast = km1,
                       fact = 1, 
                       name = 'moths_1km_effort')

mths_10km <- get_effort(coords_df = cbind(mdf_no_spp$lon, mdf_no_spp$lat),
                        rast = km10,
                        fact = 1, 
                        name = 'moths_10km_effort')

par(mfrow = c(1, 2))
plot((mths_1km), main = 'moths, 1km effort')
plot((mths_10km), main = 'moths, 10km effort')
par(mfrow = c(1, 1))


if(Write){
  
  writeRaster(mths_1km, filename = '../../data/edited_insect_data/moth/effort_layers/moth_1km_effort_layer.grd',
              format = 'raster', overwrite = T)
  
  
  writeRaster(mths_10km, filename = '../../data/edited_insect_data/moth/effort_layers/moth_10km_effort_layer.grd',
              format = 'raster', overwrite = T)
  
  
}


```

## butterflies 1 + 10km

```{r, butterfly_1_10km}

butt_1km <- get_effort(coords_df = cbind(bdf_no_spp$lon, bdf_no_spp$lat),
                       rast = km1,
                       fact = 1, 
                       name = 'butterfly_1km_effort')

butt_10km <- get_effort(coords_df = cbind(bdf_no_spp$lon, bdf_no_spp$lat),
                        rast = km10,
                        fact = 1, 
                        name = 'butterfly_10km_effort')

par(mfrow = c(1,2))
plot((butt_1km), main = 'butterfly, 1km effort')
plot((butt_10km), main = 'butterfly, 10km effort')
par(mfrow = c(1,1))

if(Write){
  
  writeRaster(butt_1km, filename = '../../data/edited_insect_data/butterfly/effort_layers/butterfly_1km_effort_layer.grd',
              format = 'raster', overwrite = T)
  
  
  writeRaster(butt_10km, filename = '../../data/edited_insect_data/butterfly/effort_layers/butterfly_10km_effort_layer.grd',
              format = 'raster', overwrite = T)
  
  
}


length(unique(bdf$yr)) * 366 * 10000 # times the number of cells within the block of interest. I.e. in a 10km grid cell there are 100 * 100m grids = 10 000

```


## Rank-abundance curves

To do this for each dataset, we first need to thin the data to only have one value per grid cell (i.e. removing duplicates of the same species in the same grid cell through time).The function counts the number of data points that fall within a given cell using the get_effort() function above which returns a raster of counts. This raster is converted to a data frame of 1s and 0s where counts > 0. Returns a data frame with species, the sum of all cells where counts>0 (i.e. number of cells with anything in) and the total number of cells as a checking mechanism.

```{r, get_prevalence_function, message=FALSE}

## function to get the prevalence of each species in a grid of any size
## it takes any species data, a raster and a vector (or list)
## of species names
## The function counts the number of data points that fall within a given cell
## using the get_effort() function above which returns a raster of counts.
## This raster is converted to a data frame of 1s and 0s where counts > 0.
## Returns a data frame with species, the sum of all cells where
## counts>0 (i.e. number of cells with anything in) and the total number of
## cells as a checking mechanism.
get_prevalence <- function(sp_data, res_rast, sp_vec){
  
  out <- lapply(1:length(sp_vec), FUN = function(s){
    
    sp_dat <- sp_data[sp_data$sp_n == sp_nam[s],]
    
    sp_rast <- get_effort(coords_df = cbind(sp_dat$lon, sp_dat$lat),
                          rast = res_rast, # 1km raster 
                          fact = 1, 
                          name = NULL) 
    
    r_v <- raster::as.data.frame(sp_rast) %>% 
      mutate(n = ifelse(layer>0, 1,0),
             tot_cells = length(layer)) %>% 
      na.omit()
    
    sp_1km <- data.frame(species = sp_nam[s],
                         counts = sum(r_v$n, na.omit = T),
                         tot_cells = unique(r_v$tot_cells))
    
    
  })
  
  return(do.call(rbind, out))
  
}



```


```{r, get_prevalence, message=F}

# moths
sp_nam <- unique(mdf$sp_n)

m_prev <- get_prevalence(sp_data=mdf, res_rast = km1, sp_vec = sp_nam)


# butterflies
sp_nam <- unique(bdf$sp_n)

b_prev <- get_prevalence(sp_data=bdf, res_rast = km1, sp_vec = sp_nam)


```


Plot the outputs

```{r, rank_abund_plots}

# moths
m_prev %>% 
  ggplot(aes(x = reorder(species, -counts), y = counts)) +
  geom_point() +
  labs(title = 'Moths, 1km prevalence', y='cell counts') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# butterflies
b_prev %>% 
  ggplot(aes(x = reorder(species, -counts), y = counts)) +
  geom_point() +
  labs(title = 'Butterflies, 1km prevalence', y='cell counts') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```










