---
title: "Making effort layers"
output: html_notebook
---


```{r}

library(raster)
library(tidyverse)

Write = TRUE

```


For the simulation work, we need to make an effort layer of recording activities for moths and butterflies. This is going to be the number of unique days with a sighting of any species per grid cell.

## moths

```{r, load_data}

# moth
mdf <- read_csv('../../data/edited_insect_data/moth/DayFlyingMoths_EastNorths_no_duplicates.csv')

# butterflies
bdf <- read_csv('../../data/edited_insect_data/butterfly/butterfly_EastNorths_no_duplicates.csv')

# 100m raster
env <- raster::stack('../../data/environmental_data/edat_nocorrs_nosea_cropped.grd')
env <- subset(env, "elevation_UK")

par(mfrow = c(1,2))
plot(env, main = 'moths')
points(x = mdf$lon, y = mdf$lat, pch = 20, cex = 0.1)


plot(env, main = 'butterflies')
points(x = bdf$lon, y = bdf$lat, pch = 20, cex = 0.1)

par(mfrow = c(1,1))


```

## aggregate raster to desired resolution

```{r, agg_resolution}

km1 <- raster::aggregate(env, 10)
km10 <- raster::aggregate(env, 100)

```

## Changing grouping level

```{r}

mdf_no_spp <- mdf  %>% 
  distinct(lon, lat, date)


bdf_no_spp <- bdf  %>% 
  distinct(lon, lat, date)


```



## effort function

```{r, effort_function}

get_effort <- function(coords_df, # two column matrix of (lon, lat)
                       rast, # raster for initial filling
                       fact, # factor to increase spatial resolution of raster - relative to initial resolution
                       name = NULL) # name to give layer
{  
  
  
  if(fact < 1) stop('aggregation factor must be > 0')
  if(fact > 1) { print('###   aggregating   ###'); rast <- raster::aggregate(rast, fact = fact) }
  
  # convert raster to 0s
  r <- rast
  r[] <- 0
  
  # get counts per cell
  counts <- table(cellFromXY(r, coords_df))
  
  # fill matrix
  r[as.numeric(names(counts))] = counts
  
  print('###   masking   ###')
  agg_mask <- raster::mask(r, rast)
  names(agg_mask) <- name
  
  return(agg_mask)
}



```

## moths 1 + 10km

```{r, moths_1+10km}

mths_1km <- get_effort(coords_df = cbind(mdf_no_spp$lon, mdf_no_spp$lat),
                       rast = km1,
                       fact = 1, 
                       name = 'moths_1km_effort')

mths_10km <- get_effort(coords_df = cbind(mdf_no_spp$lon, mdf_no_spp$lat),
                        rast = km10,
                        fact = 1, 
                        name = 'moths_10km_effort')

par(mfrow = c(1, 2))
plot(log(mths_1km), main = 'moths, 1km effort')
plot(log(mths_10km), main = 'moths, 10km effort')
par(mfrow = c(1, 1))


if(Write){
  
  writeRaster(mths_1km, filename = '../../data/edited_insect_data/moth/effort_layers/moth_1km_effort_layer.grd',
              format = 'raster', overwrite = T)
  
  
  writeRaster(mths_10km, filename = '../../data/edited_insect_data/moth/effort_layers/moth_10km_effort_layer.grd',
              format = 'raster', overwrite = T)
  
  
}


```

## butterflies 1 + 10km

```{r, butterfly_1_10km}

butt_1km <- get_effort(coords_df = cbind(bdf_no_spp$lon, bdf_no_spp$lat),
                       rast = km1,
                       fact = 1, 
                       name = 'butterfly_1km_effort')

butt_10km <- get_effort(coords_df = cbind(bdf_no_spp$lon, bdf_no_spp$lat),
                        rast = km10,
                        fact = 1, 
                        name = 'butterfly_10km_effort')

par(mfrow = c(1,2))
plot(log(butt_1km), main = 'butterfly, 1km effort')
plot((butt_10km), main = 'butterfly, 10km effort')
par(mfrow = c(1,1))

if(Write){
  
  writeRaster(butt_1km, filename = '../../data/edited_insect_data/butterfly/effort_layers/butterfly_1km_effort_layer.grd',
              format = 'raster', overwrite = T)
  
  
  writeRaster(butt_10km, filename = '../../data/edited_insect_data/butterfly/effort_layers/butterfly_10km_effort_layer.grd',
              format = 'raster', overwrite = T)
  
  
}


length(unique(bdf$yr)) * 366 * 10000 # times the number of cells within the block of interest. I.e. in a 10km grid cell there are 100 * 100m grids = 10 000

```


## Rank-abundance curves

To do this for each dataset, we first need to thin the data to only have one value per grid cell (i.e. removing duplicates of the same species in the same grid cell through time).

```{r, get_abundance_function}

# moths
mdf_thin <- mdf %>% 
  mutate(thinned_id = paste(sp_n, TO_GRIDREF, lon, lat)) %>% 
  distinct(thinned_id, .keep_all = TRUE)

sp_nam <- unique(mdf_thin$sp_n)

# for(s in 1:length(sp_nam)){
#   
#   sp_dat <- mdf_thin[mdf_thin$sp_n == sp_nam[s],]
#   
#   suppressMessages(
#     p_rast <- get_effort(coords_df = cbind(sp_dat$lon, sp_dat$lat),
#                          rast = km1, # 1km raster 
#                          fact = 1, 
#                          name = NULL) )
#   
#   r_v <- raster::as.data.frame(sp_rast) %>% 
#     mutate(n = ifelse(layer>0, 1,0),
#            tot_cells = length(layer)) %>% 
#     na.omit()
#   
#   sp_1km <- data.frame(species = sp_nam[s],
#                        counts = sum(r_v$n, na.omit = T),
#                        tot_cells = unique(r_v$tot_cells))
#   
#   
# }

get_abundance <- function(sp_data, res_rast, sp_vec){
  
  out <- lapply(1:length(sp_vec), FUN = function(s){
    
    sp_dat <- sp_data[sp_data$sp_n == sp_nam[s],]
    
    sp_rast <- get_effort(coords_df = cbind(sp_dat$lon, sp_dat$lat),
                          rast = res_rast, # 1km raster 
                          fact = 1, 
                          name = NULL) 
    
    r_v <- raster::as.data.frame(sp_rast) %>% 
      mutate(n = ifelse(layer>0, 1,0),
             tot_cells = length(layer)) %>% 
      na.omit()
    
    sp_1km <- data.frame(species = sp_nam[s],
                         counts = sum(r_v$n, na.omit = T),
                         tot_cells = unique(r_v$tot_cells))
    
    
  })
  
  return(do.call(rbind, out))
  
}




# butterflies
bdf_thin <- bdf %>% 
  mutate(thinned_id = paste(sp_n, TO_GRIDREF, lon, lat)) %>% 
  distinct(thinned_id, .keep_all = TRUE)


```

The plots

```{r, rank_abund_plots}

mdf_thin %>% 
  group_by(sp_n) %>% 
  tally() %>% 
  ungroup() %>% 
  mutate(prop = n/sum(n)) %>% 
  ggplot(aes(x = reorder(sp_n, -prop), y = (prop))) +
  geom_point() +
  # geom_path() +
  # coord_flip() +
  labs(x = 'species', y = 'proportional abundance (n/sum(n))',
       title = 'moth rank abundance') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))



bdf_thin %>% 
  group_by(sp_n) %>% 
  tally() %>% 
  ungroup() %>% 
  mutate(prop = n/sum(n)) %>% 
  ggplot(aes(x = reorder(sp_n, -prop), y = (prop))) +
  geom_point() +
  # geom_path() +
  # coord_flip() +
  labs(x = 'species', y = 'proportional abundance (n/sum(n))',
       title = 'butterfly rank abundance') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


```










